---
title: Using SingleR to annotate single-cell RNA-seq data
author:
- name: Aaron Lun
- name: Jared M. Andrews
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
- name: Friederike Dündar
  affiliation: Applied Bioinformatics Core, Weill Cornell Medicine
- name: Daniel Bunis
  affiliation: Bakar Computational Health Sciences Institute, University of California San Francisco, San Francisco, CA
date: "Revised: September 3, 2019"
output:
  BiocStyle::html_document:
    toc_float: true
package: SingleR
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Annotating scRNA-seq data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Introduction

`r Biocpkg("SingleR")` is an automatic annotation method for single-cell RNA sequencing (scRNAseq) data [@aran2019reference].
Given a reference dataset of samples (single-cell or bulk) with known labels, it labels new cells from a test dataset based on similarity to the reference set.
Specifically, for each test cell:

1. We compute the Spearman correlation between its expression profile and that of each reference sample.
2. We define the per-label score as a fixed quantile (by default, 0.8) of the distribution of correlations.
3. We repeat this for all labels and we take the label with the highest score as the annotation for this cell.
4. We optionally perform a fine-tuning step:
  - The reference dataset is subsetted to only include labels with scores close to the maximum.
  - Scores are recomputed using only marker genes for the subset of labels.
  - This is iterated until one label remains.

Automatic annotation provides a convenient way of transferring biological knowledge across datasets.
In this manner, the burden of interpreting clusters and defining marker genes only has to be done once.

# Using the built-in references

`r Biocpkg("SingleR")` provides several reference datasets (mostly derived from bulk RNA-seq or microarray data) through dedicated data retrieval functions.
For example, to obtain reference data from the Human Primary Cell Atlas:

```{r}
library(SingleR)
hpca.se <- HumanPrimaryCellAtlasData()
hpca.se
```

We use this reference in the `SingleR()` function to annotate a scRNA-seq dataset from @lamanno2016molecular.
For the sake of speed, we will only label the first 100 cells from this dataset.

```{r}
library(scRNAseq)
hESCs <- LaMannoBrainData('human-es')
hESCs <- hESCs[,1:100]

# Restrict to common genes between test and reference data:
library(scater)
common <- intersect(rownames(hESCs), rownames(hpca.se))
hpca.se <- hpca.se[common,]
hESCs <- hESCs[common,]
hESCs <- logNormCounts(hESCs)

pred.hpca <- SingleR(test = hESCs, ref = hpca.se, labels = hpca.se$label.main)
pred.hpca
```

Each row of the output `DataFrame` contains prediction results for a single cell.
Labels are shown before fine-tuning (`first.labels`), after fine-tuning (`labels`) and after pruning (`pruned.labels`), along with the associated scores.
We summarize the distribution of labels across our subset of cells:
[6~
```{r}
table(pred.hpca$labels)
```

At this point, it is worth noting that `r Biocpkg("SingleR")` is workflow/package agnostic.
The above example uses `SummarizedExperiment` objects, but the same functions will accept any (log-)normalized expression matrix.

# Using single-cell references

## Setting up the data

Here, we will use two human pancreas datasets from the `r Biocpkg("scRNAseq")` package.
The aim is to use one pre-labelled dataset to annotate the other unlabelled dataset.
First, we set up the @muraro2016singlecell dataset to be our reference.

```{r}
library(scRNAseq)
sceM <- MuraroPancreasData()

# One should normally do cell-based quality control at this point, but for
# brevity's sake, we will just remove the unlabelled libraries here.
sceM <- sceM[,!is.na(sceM$label)]
table(sceM$label)

sceM <- logNormCounts(sceM)
```

We then set up our test dataset from @grun2016denovo.
To speed up this demonstration, we will subset to the first 100 cells.

```{r}
sceG <- GrunPancreasData()
sceG <- sceG[,colSums(counts(sceG)) > 0]

# Some of the cells have no spike-in counts, so we
# set use_altexps=FALSE to avoid normalization errors.
# This is fine as we don't use the spike-ins anyway.
sceG <- logNormCounts(sceG, use_altexps=FALSE)

sceG <- sceG[,1:100]
```

We then restrict to common genes:

```{r}
common <- intersect(rownames(sceM), rownames(sceG))
sceM <- sceM[common,]
sceG <- sceG[common,]
```

## Defining custom markers

The default marker definition in `SingleR()` is intended for references derived from bulk RNA-seq data.
When using single-cell data as a reference, we suggest building your own marker list.
This involves a series of pairwise comparisons between labels to define markers that distinguish each label from another,
and is easy to perform with functions from `r Biocpkg("scran")`.
For example, we can perform pairwise $t$-tests and obtain the top 10 marker genes from each pairwise comparison.

```{r}
library(scran)
out <- pairwiseTTests(logcounts(sceM), sceM$label, direction="up")
markers <- getTopMarkers(out$statistics, out$pairs, n=10)
```

We then supply these genes to `SingleR()` directly via the `genes=` argument.
A more focused gene set also allows annotation to be performed more quickly compared to the default approach.

```{r}
pred <- SingleR(test=sceG, ref=sceM, labels=sceM$label, genes=markers)
table(pred$labels)
```

In some cases, markers may only be available for specific labels rather than for pairwise comparisons between labels.
This is accommodated by supplying a named list of character vectors to `genes`.
Note that this is likely to be less powerful than the list-of-lists approach as information about pairwise differences is discarded.

```{r}
label.markers <- lapply(markers, unlist, recursive=FALSE)
pred2 <- SingleR(test=sceG, ref=sceM, labels=sceM$label, genes=label.markers)
table(pred$labels, pred2$labels)
```

# Annotation diagnostics

`r Biocpkg("SingleR")` provides a few basic yet powerful visualization tools.
`plotScoreHeatmap()` displays the scores for all cells across all reference labels,
which allows users to inspect the confidence of the predicted labels across the dataset.
We can also display clusters (or other metadata information) for each cell by setting `clusters=` or `annotation_col=`.
In this case, we display which donor the cells came from and the labels assigned to each cell.

```{r}
plotScoreHeatmap(pred, show.labels = TRUE,
    annotation_col=data.frame(donor=sceG$donor,
        row.names=rownames(pred)))
```

The `pruneScores()` function will remove potentially poor-quality or ambiguous assignments.
In particular, ambiguous assignments are identified based on the per-cell delta, i.e., the difference between the score for the assigned label and the median across all labels for each cell.
Low deltas indicate that the assignment is uncertain, which is especially relevant if the cell's true label does not exist in the reference.
The exact threshold used for pruning is identified using an outlier-based approach that accounts for differences in the scale of the correlations in various contexts.

```{r}
to.remove <- pruneScores(pred)
summary(to.remove)
```

By default, `SingleR()` will also report pruned labels where low-quality assignments are replaced with `NA`.
However, the default pruning thresholds may not be appropriate for every dataset - see `?pruneScores` for a more detailed discussion.
We also provide the `plotScoreDistribution()` to help determine whether the thresholds are appropriate.
This displays the per-label distribution of the differences-from-median across cells, from which `pruneScores()` defines an appropriate threshold as 3 median absolute deviations (MADs) below the median.

```{r}
plotScoreDistribution(pred, show = "delta.med", ncol = 3, show.nmads = 3)
```

If some tuning parameters must be adjusted, we can simply call `pruneScores()` directly with adjusted parameters.
Here, we set labels to `NA` if they are to be discarded, which is also how `SingleR()` marks such labels in `pruned.labels`.

```{r}
new.pruned <- pred$labels
new.pruned[pruneScores(pred, nmads=5)] <- NA
table(new.pruned, useNA="always")
```

# Available reference data sets

The [legacy SingleR package](https://github.com/dviraran/SingleR/tree/master/data) provides RDA files that contain normalized expression values and cell types labels based on bulk RNA-seq, microarray and single-cell RNA-seq data from:

* Blueprint [@blueprintRef] and Encode [@encodeRef],
* the Human Primary Cell Atlas [@hpcaRef],
* the murine [ImmGen](http://www.immgen.org/) [@ImmGenRef], and
* a collection of mouse data sets downloaded from GEO [@Benayoun2019].

The bulk RNA-seq and microarray data sets of the first three reference data sets were obtained from pre-sorted cell populations, i.e., the cell labels of these samples were mostly derived based on the respective sorting/purification strategy, not via *in silico* prediction methods.

Three additional reference datasets from bulk RNA-seq and microarray data for immune cells have also been prepared.
Each of these datasets were also obtained from pre-sorted cell populations:

* The [Database for Immune Cell Expression(/eQTLs/Epigenomics)](https://dice-database.org) [@diceRef],
* Novershtern Hematopoietic Cell Data - [GSE24759](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE24759) - formerly known as Differentiation Map [@dmapRef], and
* Monaco Immune Cell Data - [GSE107011](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE107011)  [@monaco_immuneRef].

The characteristics of each dataset are summarized below:

| Data retrieval |  Organism  | Samples | Sample types |  No. of main labels  | No. of fine labels | Cell type focus |
|------------------|----------|----------|-------------|----------------------|------------|----------|
|`HumanPrimaryCellAtlasData()`| human | 713 | microarrays of sorted cell populations  | 37 |  157 | Non-specific |
|`BlueprintEncodeData()` |  human | 259 | RNA-seq | 24 | 43 | Non-specific |
|`DatabaseImmuneCellExpressionData()` | human | 1561 | RNA-seq | 5 | 15 | Immune |
|`NovershternHematopoieticData()` | human | 211 | microarrays of sorted cell populations | 17 | 38 | Hematopoietic & Immune |
|`MonacoImmuneData()` | human | 114 | RNA-seq | 11 | 29 | Immune |
|`ImmGenData()`|  mouse | 830  | microarrays of sorted cell populations | 20 | 253 | Hematopoietic & Immune |
|`MouseRNAseqData()`| mouse |358  |RNA-seq| 18  | 28 | Non-specific |

Details for each dataset can be viewed on the corresponding help page (e.g. `?ImmGenData`).
The available sample types in each set can be viewed in the collapsible sections below.

<details>
  <summary>`BlueprintEncodeData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- BlueprintEncodeData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`HumanPrimaryCellAtlasData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- HumanPrimaryCellAtlasData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`DatabaseImmuneCellExpressionData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- DatabaseImmuneCellExpressionData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`NovershternHematopoieticData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- NovershternHematopoieticData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`MonacoImmuneData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- MonacoImmuneData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`ImmGenData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- ImmGenData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

<details>
  <summary>`MouseRNAseqData` Labels</summary>

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(SingleR)
ref <- MouseRNAseqData()
samples <- cbind(ref$label.main, ref$label.fine)
colnames(samples) <- c("label.main", "label.fine")

kable(unique(samples), format = "markdown")
```
</details>

# Separating training from classification

Advanced users can split the `SingleR()` workflow into two separate training and classification steps.
This means that training (e.g., marker detection, assembling of nearest-neighbor indices) only needs to be performed once.
The resulting data structures can then be re-used across multiple classifications with different test datasets, provided the test feature set is identical to or a superset of the features in the training set.
For example:

```{r}
trained <- trainSingleR(sceM, labels=sceM$label, genes=markers)
pred2b <- classifySingleR(sceG, trained)
table(pred$labels, pred2b$labels)
```

Other efficiency improvements are possible through several arguments:

- Switching to an approximate algorithm for the nearest neighbor search in `trainSingleR()` via the `BNPARAM=` argument from the `r Biocpkg("BiocNeighbors")` package.
- Parallelizing the fine-tuning step in `classifySingleR()` with the `BPPARAM=` argument from the `r Biocpkg("BiocParallel")` package.

These arguments can also be specified in the `SingleR()` command.

# Session information

```{r}
sessionInfo()
```

# References
